---
title: "S3 method for class vector in R"
Author: "Justin M. Shea"
output: 
    github_document: default
    pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## {Question Link](https://stackoverflow.com/questions/45521273/s3-method-for-class-vector-in-r)

Is it possible to create an S3 method for all vector types at once? I.e. one that is called for a numeric, integer, character etc. vector. Something like this:

```{r, eval=FALSE}

first_element <- function(x, ...){
  UseMethod("first_element", x)
}

first_element.vector <- function(x){
  x[1]
}

first_element(1:3)
```

## Answer

By type, I assume you mean `class`. The `structure` function will allow you to define multiple classes of a vector.

First, create some data and look at its class.

```{r}
set.seed(1)
vector <- rnorm(100, mean = 1, sd = 3)
class(vector)
head(vector)
```


Now use the structure function on your vector to define additional classes. Then check the class again and call str function to see the classes and data of the new_vector object. 

```{r}
new_vector <- structure(vector, class = c("character", "integer", "numeric", "double"))
class(new_vector)

str(new_vector)
```